---
title: "M05-Visualizing Data with Tables"
author: "Aldo Naranjo"
date: '`r format(Sys.time())`'
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: 2
    toc-location: right-body
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-fold: show
    code-tools: true
    code-link: true
    highlight-style: github
    df-print: paged
    embed-resources: true
editor: visual
execute:
  freeze: auto
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(gt)
library(gtExtras)
```

## Reflections on gt and gtExtras {#sec-essay}

### What I Learned from the Videos {#sec-learned}

> *Summarize what you learned from the videos you watched.*

The videos this week walked through the `gt` package from a few different angles, and by the end of it I felt like I had a pretty solid picture of what it can do and why it matters.

The first video, with Rich Iannone (one of the package authors), showed how the package works iteratively. You start with a basic table and keep piping in additional functions until it looks the way you want. He walked through `tab_header()` for titles, `cols_label()` for renaming columns, `fmt_number()` and `fmt_currency()` for formatting values, `cols_width()` for controlling column widths, `tab_source_note()` for citations, and `tab_footnote()` for attaching contextual notes to specific cells. One thing that stuck with me was how gt automatically handles footnote ordering from top to bottom, left to right, so you never have to manage that numbering yourself. He also showed `tab_style()` for cell-level styling and `data_color()` for adding color scales to numeric columns, which is basically a heat map embedded directly in the table.

The second video approached it more from a teaching angle, building a table step by step using income quintile data. It reinforced the same core functions but added something useful: the idea of saving the table as an object and then continuing to modify it across multiple code chunks, similar to how you would layer onto a `ggplot` object. He also covered `opt_table_font()` for changing fonts, `cols_align()` for centering numeric columns, and `tab_options()` for fine-tuning borders and divider lines.

The third video was a conversation with the creator of `gtExtras`, and it got into best practices for table design more broadly, drawing on John Schwabish's guidelines. The main takeaways there were about intentional design choices: offsetting headers from the body, using subtle dividers instead of a grid of lines everywhere, right-aligning numbers, left-aligning text, grouping similar data with `group_by()`, and choosing an appropriate level of precision for decimal places. The `text_transform()` function was also shown as a way to apply any R logic to cell content, which is extremely flexible.

The fourth video introduced `gtExtras` specifically, showing how you can embed inline graphics like distribution plots, sparklines, and bar charts directly into table cells using `gt_plt_dist()`, `gt_plt_sparkline()`, and similar functions. It also showed the one-line theme functions like `gt_theme_guardian()` and `gt_theme_538()`, which apply professional data journalism styling in a single call rather than dozens of lines of `tab_options()` and `tab_style()`.

### What I Liked About gt and gtExtras {#sec-liked}

> *What did you like about the gt and gtExtras packages demonstrated in the videos?*

A few things stood out. The pipe-based workflow feels natural if you already work in the tidyverse. You can wrangle your data with `dplyr`, hand it off to `gt()`, and keep adding layers without breaking the flow of the code. That iterative, additive approach makes it easy to build incrementally and catch mistakes early.

I also really liked the footnote system. The fact that gt indexes footnotes automatically based on their position in the table means you can add or remove footnotes without manually renumbering anything. That is the kind of friction-reducing detail that makes a package genuinely pleasant to use.

`gtExtras` was impressive for a different reason. The ability to embed a sparkline or distribution plot inside a table cell blurs the line between tables and charts in a really useful way. You get the precision of exact numbers alongside a visual representation of the distribution or trend in the same row. That is something neither a standalone chart nor a plain table can do on its own.

### How Tables Complement Charts {#sec-complement}

> *How do tables complement charts for data visualization?*

Charts and tables are good at different things, and they work best when you think of them as partners rather than substitutes. Charts are designed for pattern recognition: you can see a trend, a spike, or a cluster in a second. But charts sacrifice precision to do that. If someone asks you exactly what a specific value was in Q3 of 2019, the chart might give you a rough idea but not an exact number.

Tables flip that trade-off. They are built for precision and for comparison of specific values across a well-defined set of categories. The trade-off is that visual patterns are harder to extract at a glance. A table with 40 rows of numbers does not immediately communicate that one category is dramatically outperforming the others.

Used together, you can lead with a chart to establish the shape of the story and then use a table to support specific claims with exact figures. Or, with something like `gtExtras`, you can do both in the same object by embedding inline charts into table cells. That is a genuinely new kind of artifact that did not really exist before packages like this one.

### When I Would Prefer a Table Over a Chart {#sec-prefer}

> *Under what circumstances would you prefer to use tables rather than charts to visualize data?*

There are a few situations where a table is the clear better choice for me. First, when the audience needs to look up specific values rather than understand a trend. A stakeholder who wants to know the exact revenue figure for a particular region in a particular quarter is going to be frustrated by a bar chart that makes them estimate.

Second, when you are comparing multiple metrics across multiple categories simultaneously. A chart can usually only show one or two variables clearly before it gets cluttered. A well-structured table can show five or six metrics for ten categories and still be readable if the formatting is done right.

Third, in formal written reports or slides where the table will be referenced directly, like a policy brief or an academic paper. Tables carry a kind of documentary credibility that charts sometimes do not, especially when the exact numbers matter for accountability or citation purposes.

In my MSDM context, tables make a lot of sense for presenting campaign performance data across multiple channels or time periods, where someone needs to be able to read off exact figures and compare them across rows.

------------------------------------------------------------------------

## Building a Table with gt and gtExtras {#sec-table}

### About the Data {#sec-data}

> *Use the synthetic data introduced this week for the lecture. Pick some variables of your interest from the data and share some descriptive statistics using gt and gtExtras functions together to show numbers and small charts in one table.*

For this table I am using the `gtcars` dataset, which is built into the `gt` package and was featured in the official `gt` case study linked in the course materials. It contains data on 47 high-end vehicles including manufacturer, model, horsepower, torque, fuel economy, and MSRP. I am summarizing performance and price statistics by country of origin, which gives a natural grouping structure for comparison.

### Data Wrangling {#sec-wrangle}

```{r}
#| label: wrangle-gtcars
#| code-fold: true

# Summarize gtcars by country of origin
# Keeping a list column for gtExtras inline bar charts
gtcars_summary <- gtcars |>
  group_by(ctry_origin) |>
  summarize(
    n_cars       = n(),
    mean_hp      = round(mean(hp, na.rm = TRUE), 2),
    mean_mpg_c   = round(mean(mpg_c, na.rm = TRUE), 2),
    mean_msrp    = round(mean(msrp, na.rm = TRUE), 2),
    median_msrp  = round(median(msrp, na.rm = TRUE), 2),
    msrp_list    = list(msrp),
    hp_list      = list(hp)
  ) |>
  arrange(desc(mean_msrp))
```

### The Table {#sec-gt-table}

```{r}
#| label: tbl-gtcars-summary
#| tbl-cap: "Summary of high-performance vehicles in the gtcars dataset, grouped by country of manufacturer origin."

gtcars_summary |>
  gt(rowname_col = "ctry_origin") |>

  # Title and subtitle
  tab_header(
    title    = md("**Performance & Price by Country of Origin**"),
    subtitle = md("A summary of *47 high-end vehicles* from the `gtcars` dataset")
  ) |>

  # Stub head label
  tab_stubhead(label = md("**Country**")) |>

  # Spanner column label grouping price columns
  tab_spanner(
    label   = md("**Price (USD)**"),
    columns = c(mean_msrp, median_msrp, msrp_list)
  ) |>

  # Spanner column label grouping performance columns
  tab_spanner(
    label   = md("**Performance**"),
    columns = c(mean_hp, mean_mpg_c, hp_list)
  ) |>

  # Relabel columns
  cols_label(
    n_cars      = "# Cars",
    mean_hp     = "Avg HP",
    mean_mpg_c  = "Avg MPG (City)",
    mean_msrp   = "Mean MSRP",
    median_msrp = "Median MSRP",
    msrp_list   = "MSRP Distribution",
    hp_list     = "HP Distribution"
  ) |>

  # Format currency columns
  fmt_currency(
    columns  = c(mean_msrp, median_msrp),
    currency = "USD",
    decimals = 2
  ) |>

  # Format numeric columns to 2 decimal places
  fmt_number(
    columns  = c(mean_hp, mean_mpg_c),
    decimals = 2
  ) |>

  # Inline boxplots for MSRP distribution using gtExtras
  gt_plt_dist(
    msrp_list,
    type    = "boxplot",
    fig_dim = c(5, 28)
  ) |>

  # Inline boxplots for HP distribution using gtExtras
  gt_plt_dist(
    hp_list,
    type    = "boxplot",
    fig_dim = c(5, 28)
  ) |>

  # Color the mean MSRP column by value
  gt_color_rows(
    columns   = mean_msrp,
    palette   = c("#d4f1c0", "#1a7a29"),
    domain    = NULL
  ) |>

  # Color the avg HP column by value
  gt_color_rows(
    columns   = mean_hp,
    palette   = c("#cce5ff", "#004085"),
    domain    = NULL
  ) |>

  # Source note
  tab_source_note(
    source_note = md(
      "**Source:** `gtcars` dataset from the **gt** R package (Iannone et al.). Data originally sourced from Edmunds."
    )
  ) |>

  # Footnote on mean MSRP column label
  tab_footnote(
    footnote  = "Mean MSRP calculated across all trims listed for each country.",
    locations = cells_column_labels(columns = mean_msrp)
  ) |>

  # Footnote on HP column label
  tab_footnote(
    footnote  = "Horsepower figures reflect manufacturer-reported peak HP.",
    locations = cells_column_labels(columns = mean_hp)
  ) |>

  # Footnote on Italy row (highest MSRP)
  tab_footnote(
    footnote  = md("Italy leads all countries in mean MSRP, driven by **Ferrari** and **Lamborghini** models."),
    locations = cells_stub(rows = "Italy")
  ) |>

  # Column alignment
  cols_align(align = "center", columns = c(n_cars, mean_hp, mean_mpg_c)) |>

  # Apply gtExtras theme
  gt_theme_guardian() |>

  # Additional styling: bold the stub
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_stub()
  )
```

::: {.callout-note title="Reading the table"}
Each row represents a country of vehicle manufacturer origin. The two distribution columns show boxplots of MSRP and horsepower across all vehicles from that country. Green shading in **Mean MSRP** and blue shading in **Avg HP** scale with the value: darker means higher.
:::

------------------------------------------------------------------------

## Why This Table Belongs in My MSDM Project {#sec-relevance}

> *Explain why the table is worth adding to your MSDM project written report or presentation slides.*

My MSDM capstone project is focused on digital marketing performance analysis, and tables like this one translate directly to how I would present campaign data to stakeholders or in a written report.

The structure here, grouping rows by a categorical variable (country of origin, or in my project's case, marketing channel or audience segment) and summarizing multiple numeric metrics per group, is exactly the format that decision-makers need. They want to know the averages, but they also want a sense of the spread. The inline boxplots accomplish that without requiring a separate figure or appendix.

The currency formatting, footnotes explaining methodological details, and the source note establishing data provenance are all things that belong in a professional report and are easy to overlook when you are just trying to get numbers on a page. The `gt` package makes all of that feel like a natural part of building the table rather than an afterthought.

For a presentation slide version, I would simplify the table, drop some columns, and probably use `gt_theme_538()` for a cleaner look. But the underlying structure would be the same. A well-formatted summary table with embedded visuals communicates more in less space than a raw data dump or a standalone chart, which is exactly what you need when slide real estate is limited.

------------------------------------------------------------------------

## Appendix {#sec-appendix}

-   **Published report (GitHub Pages):** <https://naran115.github.io/M05-Table-Viz/Naranjo,%20Aldo-M05-Table-Viz.html>
-   **GitHub Repository:** <https://github.com/naran115/M05-Table-Viz>

::: {.callout-note title="Note on appendix links"}
Update these URLs after creating the repo and enabling GitHub Pages. The file path in the Pages link should match the exact rendered HTML filename.
:::
